NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-31 paper/Chapters/Chapter2.tex
paper modified accordingly

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a list of discrete responses for our variable of interest, the energy consumed within each time slot over the four days of survey for each participant were calculated. The percentages of total energy intake contributed by carbohydrate within each time slot were then calculated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to these percentages of the energy intake at a cut-off value of 50\%, i.e. if within a time slot where there any energy intake occurred, carbohydrate consumption was categorised by whether its energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary. Each data point included one of the following responses:

@@

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability would agree to a nurse visit/provide a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit or who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

@@

Blood samples were stored at 4 $^\circ$C, and sent directly by post to the Department of Haematology and Department of Clinical Biochemistry and Immunology, Addenbrooke's Hospital, Cambridge within two hours of their collection. Serum samples were obtained by centrifugation of the coagulated blood sample. Serum total, High-Density Lipoprotein (HDL) and Low-Density Lipoprotein (LDL) cholesterol, triglycerides (TG), fasting blood glucose, haemoglobin A1C were measured. A1C value of 6.5\% was used as the cut off point for diagnosing diabetes.

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of the $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for the outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ is the predictors for day-level (weekdays or weekends) and individual-level, such as age, and sex.
\end{itemize}

@@

If the day-level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

\begin{equation}
%\begin{aligned}
P(U_{ij1} = s_1, U_{ij2} = s_2, \cdots, U_{ijk} = s_{k}) = \sum_{t=1}^{T}P(G_{ij}=t)\prod_{k=1}^{K}P(U_{ijk} = s_k | G_{ij} = t)
%\end{aligned}
\label{MLCA}
\end{equation}
\vspace{-0.8cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of carbohydrate eating (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on the $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on the $i$th day of the survey, the total number of day-level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

@@

Person-level point estimates and 95\% confidence intervals (CIs) were determined by applying individual, nurse visiting, and blood sample weights accordingly which account for the probability of participant selection and the clustered survey design. Descriptive statistics for sample characteristics were presented as weighted means (95\% CI) or weighted percentages (95\%CI). After examining the distribution of the data, the following variables were log-transformed to improve normality: fasting blood glucose, A1C, TC, LDL, HDL, TG, and average physical activity duration per day. Weighted geometric means (95\% CI) were used for all log-transformed variables.

To see whether there is any temporal pattern for food intake eating could also be defined at the individual-level, weighted estimates of nutrients consumption across the 7-time slots of the day were calculated for each individual-level latent class. Contributions (\%) of the average energy intake within time slots were evaluated by determining the percentages of energy coming from carbohydrate, fat, protein, and alcohol intake.

For continuous variables, the \textit{F} test was used to determine differences between latent classes with Bonferroni correction to account for multiple testing across $>$ 2 classes when applicable. For categorical variables, differences between latent classes were assessed using the adjusted Pearson $\chi^2$ test for survey data.\vspace{-0.5cm}

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Covariates that are strongly related with the outcomes but not associate with carbohydrate eating groups may reduce the standard errors and improve the precisions, so they were also considered in the linear regression models. 
	\item Confounding and/or interaction effects from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test for whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated as soon as we weight the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available either. Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if any) were checked by removing them and refitting the models as a sensitivity analysis.
	\item Since under survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}

>>>>

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a list of discrete responses for our variable of interest, the energy consumed within each time slot over the four days of survey for each participant were calculated. The percentages of total energy intake contributed by carbohydrate within each time slot were then calculated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to these percentages of the energy intake at a cut-off value of 50\%, i.e. if within a time slot where any energy intake occurred, carbohydrate consumption was categorised by whether its energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary. Each data point included one of the following responses:

@@

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those who completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability of agreeing to a nurse visit/providing a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit or who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

@@

Blood samples were stored at 4 $^\circ$C, and sent directly by post to the Department of Haematology and Department of Clinical Biochemistry and Immunology, Addenbrooke's Hospital, Cambridge within two hours of their collection. Serum samples were obtained by centrifugation of the coagulated blood sample. Serum total, High-Density Lipoprotein (HDL) and Low-Density Lipoprotein (LDL) cholesterol, triglycerides (TG), fasting blood glucose, haemoglobin HbHbA1C were measured. HbA1C value of 6.5\% was used as the cut off point for diagnosing diabetes.

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of the $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for the outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ are the predictors for respectively day-level (weekdays or weekends) and individual-level, such as age, and sex.
\end{itemize}

@@

If the day-level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators as indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

\begin{equation}
%\begin{aligned}
P(U_{ij1} = s_1, U_{ij2} = s_2, \cdots, U_{ijk} = s_{k}) = \sum_{t=1}^{T}P(G_{ij}=t)\prod_{k=1}^{K}P(\mathbf{U_{ijk} = s_k} | G_{ij} = t)
%\end{aligned}
\label{MLCA}
\end{equation}
\vspace{-0.8cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of carbohydrate eating (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on the $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on the $i$th day of the survey, the total number of day-level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(\mathbf{U_{ijk} = s_k}|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

@@

Person-level point estimates and 95\% confidence intervals (CIs) were determined by applying individual, nurse visiting, and blood sample weights and the clustered survey design. Descriptive statistics for sample characteristics were presented as weighted means (95\% CI) or weighted percentages (95\%CI). After examining the distribution of the data, the following variables were log-transformed to improve normality: fasting blood glucose, HbA1C, TC, LDL, HDL, TG, and average physical activity duration per day. Weighted geometric means (95\% CI) were used for all log-transformed variables.

To see whether there is any temporal pattern for food intake at the individual-level, weighted estimates of nutrients consumption across the 7-time slots of the day were calculated for each individual-level latent class. Contributions (\%) of the average energy intake within time slots were evaluated by determining the percentages of energy coming from carbohydrate, fat, protein, and alcohol intake.

For continuous variables, the \textit{F} test was used to determine differences between latent classes with Bonferroni correction to account for multiple comparisons across $>$ 2 classes when applicable. For categorical variables, differences between latent classes were assessed using the adjusted Pearson $\chi^2$ test for survey data.\vspace{-0.5cm}

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Covariates that are strongly related with the outcomes but not associate with carbohydrate eating groups may reduce the standard errors and improve the precisions, so they were also considered in the linear regression models. 
	\item Confounding and/or interaction effects from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test for whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using \textit{F}-adjusted mean residual test \parencite{archer2001goodness,archer2007goodness} which is implemented in Stata using command \code{svylogitgof} \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated because of the clustered structure of the data. Therefore, inference from such models is based on design rather than the model.  
	General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are unneeded (also not available). Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if any) were checked by removing them and refitting the models as a sensitivity analysis.
	\item Since under complex survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was used to decide whether quadratic and cubic terms of continuous variables were necessary in improving fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}
